|

### 2022-05-28 article/article.Rmd
final touches

In this paper, we propose a theory based on executive leadership and party
politics to explain how city council size impacts local public goods provision.
We incorporate partisanship and mayoral first-move advantage to a
\citet{baron1989bargaining} bargaining model. Mayors make the initial policy
proposal, which consists of investments in public goods provision, rents for
personal gains, and a vector of side-transfers for councilors. If mayors fail
to secure majoritarian support from councilors, a reversal stage occurs, in
which the decision leaves the mayor's office and councilors propose the public
investments. In this bargaining process, costs come from the side-transfers
required to secure a majority among councilors. We show that in a model without
partisan concerns, bargaining costs always increase with larger city councils,
which in turn lower public goods provision and rents. However, partisanship
decreases bargaining costs when the chances of electing government-aligned
councilors are sufficiently high. Lower bargaining costs release funds for
mayors to invest in public goods. In this case, a larger city council may add
more members to the government coalition, thus reducing the mayor's bargaining
costs and increasing the equilibrium public goods provision.

@@

To provide further evidence for our mechanisms, we surveyed 174 former
councilors who served during the 2005-2008 term, the period we analyze in this
paper. They confirm that mayors indeed use bureaucratic appointments and
political favors to secure legislative support. As our formal model shows,
bargaining with councilors requires offering transfers to the representatives.
Councilors also demand several job appointments, which are regarded as an
effective means to boost their reelection prospects and improve municipal
welfare. These findings are in line with quantitative results from a novel
dataset we built with 346,553 laws passed in 63 municipalities within ten
thousand inhabitants away from the cutoffs. While all municipalities mention
public goods in their legislation, those with larger city councils have 15
percent more proposals about public service delivery.

@@

This paper contributes to four strands of the literature. First, this paper
dialogues with a large literature on multilateral bargaining. Repeated models
of multilateral bargaining started with \citet{baron1989bargaining}, and
scholars have recently analysed how endogenous recognition power, the size of
the required majority, and partisanship impact these results
\citep{agranov2020persistence, baranski2020determinants,
calvert2005legislative, choate2019partisan}. Here we include the mayor's first
move and political partisanship to Baron and Ferejohn's
(\citeyear{baron1989bargaining}) and investigate the often-overlooked effect of
legislature size on bargaining equilibrium strategies.

@@

Finally, this paper also contributes to the literature on the causes and effects of
legislature sizes. While most papers on legislature size analyse the trade-off
between representation and efficiency \citep{frederick2008people,
jacobs2015explaining, Stigler1976}, recent articles have
explored the effects of legislature size on lawmaking and partisanship in
further detail. \citet{kirkland2014chamber} shows that larger legislatures make
it harder for politicians to know and collaborate with each other, so they rely
more heavily on party cues and committee structures.
\citet{jacobs2015explaining} show that population shifts tend to increase the
number of local legislators, but legislature size tends to decrease during
economic crises. \citet{marland2019fewer} questions the latter mechanism,
showing that political elites shrink legislatures mostly to signal to the
public the priority of their austerity agenda. In this article, we show that
council size affects welfare by changing the partisan composition of the
legislatures.

@@

The literature on legislature size highlights that the number of legislators
causes tension between representation and efficiency
\citep{frederick2008people, jacobs2015explaining, taagepera1972size}. On the
one hand, larger legislatures improve representation and facilitate the
electoral success of ethnic and political minorities
\citep{allenrepresentation, gerring2020population}. On the other hand, better
representation is often countered by the decline in efficiency in larger
legislatures \citep{Weingastetal1981}. In large councils, it is harder for
members to collaborate with each other without some type of partisanship or
formal legislative structure \citep{kirkland2014chamber}. Larger chambers also
have more potential veto players, who may decrease policymaking efficiency
\citep{tsebelis2002veto}. In this sense, balancing representation and
efficiency is key to find the optimal legislative composition
\citep{auriol2012optimal, taagepera1972size}.

@@

The city council votes the mayor's proposal and the proposal is implemented if
most councilors accept it. Otherwise, a reversal policy takes place. Reversal
policy in this context means that the mayor's proposal failed to secure the
majority of council votes. The decision process then leaves the mayor's office
and is transferred to the council. The types of reversal policies are crucial
for our argument, as they change the relative strength of the council when
legislators bargain with the mayor. Here we investigate two types of reversal
policies: A baseline reversal with no parties and a hybrid mechanism with
partisan and non-partisan concerns. The differences between these two
mechanisms are that the non-partisan councilors only care about rents, while
partisan councilors care about both rents and policy.

The mayor must convince at least half of the city councilors to support her
policy proposals. Similar to \citet{buenodemesquita2005logic}, we assume that
the selectorate, that is, the individuals who have a say in the mayor's policy
proposals, are the $N$ city councilors. The winning coalition, which is the
minimum number of members that must be convinced to implement the policy, is
equal to half of the selectorate ($N/2$). City councilors are motivated by
policy concerns $p$ and transfers received from the mayor or the councilor
nominated as the proposer \textbf{x}. We leave the definition of these
transfers open, as this allows us to analyze how mayors combine transfers with
other incentives to pressure city councilors. Examples of transfers are
granting electoral favors, funding clientelistic brokers, spending in areas in
which the councilor has political or personal interests, or providing portfolio
(municipal offices) and public jobs within the bureaucracy. The use
of transfers to influence the councilors' choices generate a bargaining cost of
$C_G$. Bargaining costs vary according to the reversal mechanism and the size
of the legislature.

Finally, after learning the bargaining costs, the mayor provides a level of
public goods provision $g$ and pockets rents $r$. Public goods provision helps
the mayor to get reelected. Rents are for the mayor's direct consumption and do
not contribute to her electoral success. This makes the mayor's expected utility a
sum of the gains from rents and the benefits from reelection. Both the
utilities from rents $u(.)$ and the probability of reelection $\pi(.)$ are
concave functions, meaning that more rents or public goods increase the utility
at a decreasing rate. The probability of reelection is multiplied by the
benefit from holding office $B_M > 0$. This benefit captures the tangible and
intangible gains that the mayor receives from holding the public office. We
assume that the benefits are high enough to rule out an equilibrium with only
rents and transfers, which is equivalent to saying that mayors strongly prefer
to be reelected. If the mayor's policy proposal is approved, then the expected
utility for the mayor is:

@@

We now consider a situation that combines both partisan and non-partisan
legislature concerns. Each councilor has a party affiliation. Party
affiliations are mutually exclusive, and if a city councilor is aligned with
the mayor's party, then she belongs to the government coalition $G \subset \{1,
2, \cdots, N \}$. Otherwise, the city councilor belongs to the opposition $O =
\{1, 2, \cdots, N \} \setminus G$. The mayor's policy generates a political
value of $p > 0$ for the councilors, and whether it adds or subtracts from the
councilor's utility depends on the party affiliation of the councilor. If
councilors reject the mayor's proposal, no public goods are provided ($p=0$).
Moreover, in equilibrium, we assume that city councilors would prefer to
receive transfers instead of providing public goods.

The solution concept we use in this model is the subgame perfect Nash
equilibrium. This game requires that the strategies follow a Nash equilibrium
in each subgame. We find the solution using backward induction. Since the model
may have infinitely repeated proposals, any strategy can be shown to be a Nash
equilibrium. Thus, a natural assumption is to extend the equilibrium concept to
require the equilibrium to be stationary. A stationary subgame perfect
equilibrium requires that, at each given point of the game, if a politician
accepts an offer at time $k+1$, she should take the same offer at time $k$.
Stationarity gives us a method to find a proposal that would be accepted at any
stage of the game. Hence, we can characterize a sequence of offers for each
point $k$ in time. For the mayor, the optimal offer is at $k=0$, representing a
no delay in policy implementation. We assume that players have no dominated
strategies, even when there is insufficient support for the policy.

@@

The optimal level of public goods provision equates the marginal gains accrued
from rents with the marginal benefits from reelection.

@@

All proofs are in section A.1 of the Supplemental Materials. The proof's
intuition involves finding what happens to the mayor's marginal utility when we
increase both $g$ and $N$. As $N$ increases discretely on a complete lattice,
we use monotone comparative statics to derive the conditions for increasing
differences \citep{ashworth2006monotone}. Increasing differences mean that
higher $N$ leads to higher equilibrium $g$.

@@

\[
x_i \geq \dfrac{1}{N}\left[ \delta^{k+1}R - \dfrac{N}{2}x_i \right] + \left[ 1 - \dfrac{1}{N} \right]\left[\dfrac{1}{2} x_i \right]
\]

We place the offer on the left-hand side. There are two components on the
right-hand side. The first is the amount that the councilor $i$ gets when she
is the proposer. It is equal to the chance that she is recognized as the
proposer times the budget in the next round minus the offers she makes to
convince half of the councilors. The second part represents the gains if the
councilor rejects the current offer but still gets an offer in the following
round. It is equal to the chance that she is not recognized as the proposer
times the chance that she receives a transfer times the transfer amount. Note
that the offer the councilor makes as the proposer is the same as the offer she
wants to receive. This is because city councilors are exchangeable. The
solution is symmetric for all councilors receiving an offer (this means that,
without loss of generality, we could have dropped the $i$ in the solution).
After some algebra, the offer $x_i$ must be greater than or equal to:

@@

The proposer always offers the minimum required to get the proposal approved.
In this case, the offer at any given stage $k$ is going to be equal to
$\underline{x}(k, N)$.

Proceeding backwards, at the mayor's proposal stage $k=0$ the mayor is going to
offer $\underline{x}(0, N) = \dfrac{2 \delta R}{2N + 1}$ to half of the
councilors. In this context, the bargaining costs for the mayor are:

@@

We decompose these costs into two components, which are the costs in terms of
rents and the costs (or benefits) from political alignment. In this context,
the bargaining costs depend on what the mayor offers for councilors in her
coalition versus what she provides to members of the opposition. As the ex-ante
chance of an opposition member being elected is equal to $1-\gamma$, after
taking the weighted averages these costs become:

@@

Also, the threshold $\underline{\gamma}$ that captures the probability of
electing a government supporter decreases along with the political value of
policy $p$ and the size of the legislature $N$. This means that intensive
partisanship and larger legislatures make it easier to satisfy the electoral
threshold.

@@

With regards to their political organization, municipalities have one mayor and
one city council. Citizens choose their representatives by direct vote in the
same election. Mayors oversee the provision of public services and propose laws
and amendments to the tax code. Proposals submitted by the mayor are subject to
the approval of the city council. Councilors, in turn, have two primary duties.
First, they should discuss and vote on legislation put forward by the mayors or
by their peers. Second, they oversee the public budget and sanction mayors who
do not comply with fiscal legislation. Councilors and mayors also provide an
array of public and personal services for their constituents, such as helping
voters access public hospitals, facilitating school enrollment, and even paying
their voters' utility bills \citep{lopez2017legisladores,
nichter2011electoral}.

>>>>

In this paper, we propose a theory based on executive leadership and party
politics to explain how city council size impacts local public goods provision.
We incorporate partisanship and mayoral first-move advantage to a
\citet{baron1989bargaining} bargaining model. Mayors make the initial policy
proposal, which consists of investments in public goods provision, rents for
personal gains, and a vector of side transfers for councilors. If mayors fail
to secure majoritarian support from councilors, a reversal stage occurs, in
which the decision leaves the mayor's office and councilors propose the public
investments. In this bargaining process, costs come from the side transfers
required to secure a majority among councilors. We show that in a model without
partisan concerns, bargaining costs always increase with larger city councils,
which in turn lower public goods provision and rents. However, partisanship
decreases bargaining costs when the chances of electing government-aligned
councilors are sufficiently high. Lower bargaining costs release funds for
mayors to invest in public goods. In this case, a larger city council may add
more members to the government coalition, thus reducing the mayor's bargaining
costs and increasing the equilibrium public goods provision.

@@

To provide further evidence for our mechanisms, we surveyed 174 former
councilors who served during the 2005-2008 term, the period we analyze in this
paper. They confirm that mayors indeed use bureaucratic appointments and
political favors to secure legislative support. As our formal model shows,
bargaining with councilors requires offering transfers to the representatives.
Councilors also demand several job appointments, which are regarded as an
effective means to boost their reelection prospects and improve municipal
welfare. These findings are in line with quantitative results from a novel
dataset we built with 346,553 laws passed in 63 municipalities within ten
thousand inhabitants away from the cutoffs. While all municipalities mention
public goods in their legislation, those with larger city councils had 15
percent more proposals about public service delivery.

@@

This paper contributes to four strands of the literature. First, this paper
dialogues with a large literature on multilateral bargaining. Repeated models
of multilateral bargaining started with \citet{baron1989bargaining}, and
scholars have recently analyzed how endogenous recognition power, the size of
the required majority, and partisanship impact these results
\citep{agranov2020persistence, baranski2020determinants,
calvert2005legislative, choate2019partisan}. Here we include the mayor's first
move advantage and political partisanship to Baron and Ferejohn
(\citeyear{baron1989bargaining}) and investigate the often-overlooked effect of
legislature size on bargaining equilibrium strategies.

@@

Finally, this paper also contributes to the literature on the causes and effects of
legislature sizes. While most papers on legislature size analyze the trade-off
between representation and efficiency \citep{frederick2008people,
jacobs2015explaining, Stigler1976}, recent articles have
explored the effects of legislature size on lawmaking and partisanship in
further detail. \citet{kirkland2014chamber} shows that larger legislatures make
it harder for politicians to know and collaborate with each other, so they rely
more heavily on party cues and committee structures.
\citet{jacobs2015explaining} show that population shifts tend to increase the
number of local legislators, but legislature size tends to decrease during
economic crises. \citet{marland2019fewer} questions the latter mechanism,
showing that political elites shrink legislatures mostly to signal to the
public the priority of their austerity agenda. In this article, we show that
council size affects welfare by changing the partisan composition of the
legislatures.

@@

The literature on legislature size highlights that the number of legislators
causes tension between representation and efficiency
\citep{frederick2008people, jacobs2015explaining, taagepera1972size}. On the
one hand, larger legislatures improve representation and facilitate the
electoral success of ethnic and political minorities
\citep{allenrepresentation, gerring2020population}. On the other hand, better
representation is often countered by the decline in efficiency in bigger
legislatures \citep{Weingastetal1981}. In large councils, it is harder for
members to collaborate with each other without some type of partisanship or
formal legislative structure \citep{kirkland2014chamber}. Larger chambers also
have more potential veto players, who may decrease policymaking efficiency
\citep{tsebelis2002veto}. In this sense, balancing representation and
efficiency is key to find the optimal legislative composition
\citep{auriol2012optimal, taagepera1972size}.

@@

The city council votes the mayor's proposal and the proposal is implemented if
councilors accept it. Otherwise, a reversal policy takes place. Reversal policy
in this context means that the mayor's proposal failed to secure the majority
of council votes. The decision process then leaves the mayor's office and is
transferred to the council. The types of reversal policies are crucial for our
argument, as they change the relative strength of the council when legislators
bargain with the mayor. Here we investigate two types of reversal policies: A
baseline reversal with no parties and a hybrid mechanism with partisan and
non-partisan concerns. The differences between these two mechanisms are that
the non-partisan councilors only care about rents, while partisan councilors
care about both rents and policy.

The mayor must convince at least half of the city councilors to support her
policy proposals. Similar to \citet{buenodemesquita2005logic}, we assume that
the selectorate, that is, the individuals who have a say in the mayor's policy
proposals, are the $N$ city councilors. The winning coalition, which is the
minimum number of members that must be convinced to implement the policy, is
equal to half of the selectorate. City councilors are motivated by policy
concerns $p$ and transfers \textbf{x} received from the mayor or the councilor
nominated as the proposer. We leave the definition of these transfers open, as
this allows us to analyze how mayors combine transfers with other incentives to
pressure city councilors. Examples of transfers are granting electoral favors,
funding clientelistic brokers, spending in areas in which the councilor has
political or personal interests, or providing portfolio (municipal offices) and
public jobs within the bureaucracy. The use of transfers to influence the
councilors' choices generate a bargaining cost of $C_G$. Bargaining costs vary
according to the reversal mechanism and the size of the legislature.

Finally, after learning the bargaining costs, the mayor offers a level of
public goods provision $g$ and pockets rents $r$. Public goods provision helps
the mayor to get reelected. Rents are for the mayor's direct consumption and do
not contribute to her electoral success. This makes the mayor's expected
utility a sum of the gains from rents and the benefits from reelection ($B_M$).
Both the utilities from rents $u(.)$ and the probability of reelection $\pi(.)$
are concave functions, meaning that more rents or public goods increase the
utility at a decreasing rate. The probability of reelection is multiplied by
the benefit from holding office. This benefit captures the tangible and
intangible gains that the mayor receives from holding the public office. We
assume that the benefits are high enough to rule out an equilibrium with only
rents and transfers, which is equivalent to saying that mayors strongly prefer
to be reelected. If the mayor's policy proposal is approved, then the expected
utility for the mayor is:

@@

We then consider a situation that combines both partisan and non-partisan
legislature concerns. Each councilor has a party affiliation. Party
affiliations are mutually exclusive, and if a city councilor is aligned with
the mayor's party, then he belongs to the government coalition $G \subset \{1,
2, \cdots, N \}$. Otherwise, the city councilor belongs to the opposition $O =
\{1, 2, \cdots, N \} \setminus G$. The mayor's policy generates a political
value of $p > 0$ for the councilors, and whether it adds or subtracts from the
councilor's utility depends on the party affiliation of the councilor. If
councilors reject the mayor's proposal, there are no partisan alignment
benefits ($p=0$). Moreover, in equilibrium, we assume that city councilors
would prefer to receive transfers instead of providing public goods.

The solution concept we use in this model is the stationary subgame perfect
Nash equilibrium. This solution concept requires that the strategies follow a
Nash equilibrium in each subgame. Since the model may have infinitely repeated
proposals, any strategy can be shown to be a Nash equilibrium. Thus, we solve
for a stationary equilibrium. A stationary subgame perfect equilibrium requires
that, at each given point of the game, if a politician accepts an offer at time
$k+1$, he should take the same offer at time $k$. Stationarity gives us a
method to find a proposal that would be accepted at any stage of the game.
Hence, we can characterize a sequence of offers for each point $k$ in time. For
the mayor, the optimal offer is at $k=0$, representing a no delay in policy
implementation. We assume that councilors play no dominated strategies, even
when there is insufficient support for the policy.

@@

@@

All proofs are in section A.1 of the Supplemental Materials. The proof's
intuition involves finding what happens to the mayor's marginal utility when we
increase both $g$ and $N$. As $N$ increases discretely, we use monotone
comparative statics to derive the conditions for increasing differences
\citep{ashworth2006monotone}. Increasing differences mean that higher $N$ leads
to higher equilibrium $g$.

@@

\[
x_i \geq \dfrac{1}{N}\left[ \delta^{k+1}R - \dfrac{N}{2}x_i \right] + \left[ 1 - \dfrac{1}{N} \right]\dfrac{x_i}{2}  
\]

We place the offer on the left-hand side. There are two components on the
right-hand side. The first is the amount that councilor $i$ gets when he is the
proposer. It is the chance that he is recognized as the proposer times the
budget in the next round minus the offers he makes to convince half of the
councilors. The second part represents the gains if the councilor rejects the
current offer but still gets an offer in the following round. It is equal to
the chance that he is not recognized as the proposer times the chance that he
receives a transfer times the transfer's amount. Note that the offer the
councilor makes as the proposer is the same as the offer he wants to receive.
This is because city councilors are exchangeable. The solution is symmetric for
all councilors receiving an offer (this means that, without loss of generality,
we could have dropped $i$ in the solution). After some algebra, the offer $x_i$
must be greater than or equal to:

@@

The proposer always offers the minimum required to get the offer approved.
In this case, the proposal at any given stage $k$ is going to be equal to
$\underline{x}(k, N)$.

Proceeding backwards, at the mayor's proposal stage $k=0$ she is going to
offer $\underline{x}(0, N) = \dfrac{2 \delta R}{2N + 1}$ to half of the
councilors. In this context, the bargaining costs for the mayor are:

@@

We decompose these costs into two components, which are the costs in terms of
rents and the costs (or benefits) from political alignment. In this context,
the bargaining costs depend on what the mayor offers for councilors in her
coalition versus what she provides to members of the opposition. As the chance
of electing a government member is equal to $\gamma$, after taking the weighted
averages, these costs become:

@@

Also, the threshold $\underline{\gamma}$ that captures the probability of
electing a government supporter decreases along with the political value of
policy $p$ and the size of the legislature $N$. This means that intensive
partisanship and larger legislatures make it easier to satisfy this electoral
threshold.

@@

With regards to their political organization, municipalities have one mayor and
one city council. Citizens choose their representatives by direct vote in the
same election. Mayors oversee the provision of public services and propose laws
and amendments to the tax code. Proposals submitted by the mayor are subject to
the approval of the city council. Councilors, in turn, have two primary duties.
First, they should discuss and vote on legislation put forward by the mayors or
their peers. Second, they oversee the public budget and sanction mayors who
do not comply with fiscal legislation. Councilors and mayors also provide an
array of public and personal services for their constituents, such as helping
voters access public hospitals, facilitating school enrollment, and even paying
their voters' utility bills \citep{lopez2017legisladores,
nichter2011electoral}.

### 2022-05-28 memorandum/memo.Rmd
final touches

Below we discuss in more detail these changes and the several others we have
made to the manuscript in response to comments from the Editors and Reviewers.
We thank the Editors again for the opportunity to revise the manuscript, and
the Reviewers for their extremely helpful comments.

@@

We thank Reviewer 1 for the opportunity to clarify why we have selected these
two sets of variables. We included an explanation for our decision on pages
14-15: "\textit{Our focus is on the impact of legislature size on education and
health care, which are the two most important public services the municipal
government should provide to citizens. According to the Brazilian Constitution,
municipalities must spend 25\% of their budget on education and 15\% on health
care. Consequently, the Ministry of Health and the Ministry of Education
provide readily available, fine-grained data on health care and education at
the municipal level, which we leverage in this study}."

@@

To reflect these pieces of information, we have rewritten paragraph five in
Section 5 of the manuscript (pages 16-17). It now reads:

>>>>

Below we discuss in more detail these changes and the several others we have
made to the manuscript in response to comments from the Editors, the Reviewers,
and the Editorial Office. We thank the Editors again for the opportunity to
revise the manuscript, and the Reviewers for their extremely helpful comments.

@@

We thank Reviewer 1 for the opportunity to clarify why we have selected these
two sets of variables. We included an explanation for our decision on page 14:
"\textit{Our focus is on the impact of legislature size on education and health
care, which are the two most important public services the municipal government
should provide to citizens. According to the Brazilian Constitution,
municipalities must spend 25\% of their budget on education and 15\% on health
care. Consequently, the Ministry of Health and the Ministry of Education
provide readily available, fine-grained data on health care and education at
the municipal level, which we leverage in this study}."

@@

To reflect these pieces of information, we have rewritten paragraph five in
Section 5 of the manuscript (page 16). It now reads:
